On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-03-29 manuscripts/PNAS/chazthm_pnas.tex
more edits

Since $f$ is a real analytic function, one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set of (Borel) measure zero. Hence, almost every $n \times m$ real matrix $\mathbf{A}$ satisfies \eqref{SparkCondition} provided \eqref{CScondition} holds for a value of $\gamma$ for some distribution. We remark that the precise relationship between $m$, $n$, and $k$ guaranteeing that $f$ is not identically zero is a challenging problem in real algebraic geometry. In any case, we set $\gamma_0$ to be the smallest known such $\gamma$.

@@

In fact, in the case $k=1$ we can relax our assumptions even further and assume that the matrix $\mathbf{B}$ has only at least as many columns as $\mathbf{A}$; as we will see, the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} then hold for some $n \times m$ submatrix of $\mathbf{B}$.  [*** ??? ***]

\begin{proof}[Proof ($k=1$)]
Note that the only 1-uniform hypergraph satisfying the singleton intersection property is $[m]$; hence, we have $\mathbf{x}_i = c_i \mathbf{e}_i$ for $c_i \in \mathbb{R} \setminus \{\mathbf{0}\}$, $i \in [m]$. In this case we may take $C_1$ to be any value satisfying $C_1^{-1} \leq \min_{j \in [m]} |c_{j}|$.

Fix $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfying \eqref{SparkCondition} and suppose that for some $\mathbf{B} \in \mathbb{R}^{n \times m'}$ ($m' \geq m$) and $1$-sparse $\mathbf{\hat x}_i \in \mathbb{R}^{m'}$ we have  $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon < L_2(\mathbf{A}) / C_1$ for all $i$. Then there are $c'_1, \ldots, c'_m \in \mathbb{R}$ and $\pi: [m] \to [m']$ with:
\begin{align}\label{1D}
\|c_j\mathbf{A}_j - c'_j\mathbf{B}_{\pi(j)}\|_2 \leq \varepsilon, \ \ \text{for all $j$}.
\end{align} 
Note that $c'_j \neq 0$ for all $j$ since otherwise by \eqref{delrho} and definition of $C_1$ we would have $\|c_j\mathbf{A}_j\|_2 < \min_{\ell \in [m]}\|c_{\ell}\mathbf{A}_{\ell}\|_2$.

We  now show that $\pi$ is injective (and thus is a permutation if $m' = m$). Suppose that $\pi(i) = \pi(j) = \ell$ for some $i \neq j$ and $\ell$. Then, $\|c_{j}\mathbf{A}_{j} - c'_{j}\mathbf{B}_{\ell}\|_2 \leq \varepsilon$ and $\|c_{i}\mathbf{A}_{i} - c'_{i} \mathbf{B}_{\ell}\|_2  \leq \varepsilon$. Scaling and summing these inequalities by $|c'_{i}|$ and $|c'_{j}|$, respectively, and applying the triangle inequality, we obtain:
\begin{align*}%\label{contra}
(|c'_{i}| + |c'_{j}|) \varepsilon
&\geq\|\mathbf{A}(c'_{i}c_{j} \mathbf{e}_{j} - c'_{j}c_{i}\mathbf{e}_{i})\|_2 \nonumber \\ 
&\geq  \left( |c'_{i}| + |c'_{j}| \right) L_2(\mathbf{A}) \min_{\ell \in [m]} |c_\ell |,
\end{align*}
%
which contradicts the upper bound $\varepsilon < L_2(\mathbf{A})/C_1$. Hence, the map $\pi$ is injective. Setting $J = \pi([m])$ and letting $P = \left( \mathbf{e}_{\pi(1)} \cdots \mathbf{e}_{\pi(m)}\right)$ and $D = \text{diag}(\frac{c'_1}{c_1},\ldots,\frac{c'_m}{c_m})$, we see that \eqref{1D} becomes, for all $j \in [m]$:
\begin{align*}%\label{k=1result}
\|\mathbf{A}_j - (\mathbf{B}_J\mathbf{PD})_j\|_2 
= \|\mathbf{A}_j - \frac{c'_j}{c_j}\mathbf{B}_{\pi(j)}\|_2 
\leq \frac{\varepsilon}{|c_j|} 
\leq C_1\varepsilon.
\end{align*}
\end{proof}

@@

\begin{definition}\label{dDef}
For any point $\mathbf{u}$, let $\text{\rm dist}(\mathbf{u}, V) := \inf \{\| \mathbf{u}-\mathbf{v} \|_2: \mathbf{v} \in V\}$ be the distance of $\mathbf{u}$ from the set $V$. For subspaces $U,V \subseteq \mathbb{R}^m$, we define: %[Kato p.197]
\begin{align}\label{d}
d(U,V) := \sup_{\mathbf{u} \in U, \|\mathbf{u}\|_2 = 1} \text{\rm dist}(\mathbf{u},V).
\end{align}
Note that $d(U,\{\textbf{0}\}) = 1$ for $U \neq \{\mathbf{0}\}$, whereas if $U = \{\textbf{0}\}$ then $d$ has no meaning; in this case, set $d(\{\textbf{0}\},V) = 0$ for any $V$.
\end{definition}
%Note that the supremum is always achievable by a point in $U$ when $U, V \subseteq \mathbb{R}^m$.

@@

Our result in combinatorial matrix analysis is the following, which we derive by the following arguments (see the Supplemental for the full proof). First, we show that the aforementioned proximity between $k$-dimensional subspaces implies a proximity between smaller subspaces spanned by columns of $\mathbf{A}$ indexed by the intersections of sets in $E$ and those spanned by as many or fewer columns of $\mathbf{B}$. Another pigeonholing argument here combined with our assumptions on $E$ (e.g., the singleton intersection property) reveals that, in fact, each column of $\mathbf{A}$ spans a subspace proximal to that spanned by some column of $\mathbf{B}$. \eqref{Cstable} is a simple consequence of this last fact, which actually constitutes the proof of the theorem for the case $k=1$. We present this special case now before restating the above arguments in greater detail.

@@

[Effective computation of stable datasets, at least in principle.]

@@

\begin{lemma}\label{MainLemma2}
Fix integers $n$ and $k < m \leq m'$ and suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ and $\mathbf{B} \in \mathbb{R}^{n \times m'}$ both satisfiy \eqref{SparkCondition}. There exists a constant $C_4 > 0$ for which for all $\varepsilon < L_2(\mathbf{A}) / C_4$ the following holds. If for some $E \subseteq2^{[m]}$ satisfying the singleton intersection property there exists a size-preserving map $\pi: E \mapsto 2^{[m']}$ satisfying:
\begin{align}\label{GapUpperBound}
d(\text{\rm Span}\{\mathbf{A}_{S}\}, \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) \leq \varepsilon \ \ \text{for all $S \in E$},
\end{align}
%
then for some $J \in {[m'] \choose m}$, permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$, we have:
\begin{align}\label{MainLemmaBPD}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq C_4 \varepsilon, \indent \text{for } j \in [m].
\end{align}
The constant $C_4$ is given by:
\begin{align}\label{Cdefm'}
C_4 := \frac{2^{|E|}}{R} \max_{j \in [m]} \|\mathbf{A}_j\|_2
\end{align}
%
with $R$ being the lesser of $\min_{\substack{F \subseteq E}} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F})$ and $\min_{\substack{F \subseteq {[m'] \choose k}}} r( \{ \text{\rm Span}\{\mathbf{B}_{S}\} \}_{S \in F})$.
\end{lemma}

\begin{proof}
The proof is very similar to the case $m' = m$, the difference being that now we may not invoke Lem.~\ref{NonEmptyLemma} (which requires $m' = m$) to infer from \eqref{fact2} that $| \cap_{S \in F(i)} \pi(S) | = 1$ for all stars $F(i)$. We circumvent this issue by instead assuming the spark condition on $\mathbf{B}$, which allows us to swap the roles of $\mathbf{A}$ and $\mathbf{B}$ in the arguments leading to the derivation of \eqref{fact2} to prove the opposite inequality for every star $F(i)$. Of course, this requires that the constant $C_4$  depend on $\mathbf{B}$. By these arguments we establish that $| \cap_{S \in F(i)} \pi(S) | = 1$ for all stars $F(i)$, yielding an injective map $\hat \pi: [m] \to [m']$. The result then follows from the proof of the case $k=1$. 
\end{proof}

@@

\begin{proof}[Proof of Thm.~\ref{DeterministicUniquenessTheorem2}]
The proof is the same as that of the case $m' = m$, only now we establish a map $\pi: E \to {[m'] \choose k}$ by pigeonholing $(k-1){m' \choose k} + 1$ vectors with respect to holes $[m']$ and eventually applying Lem.~\ref{MainLemma2} instead of Lem.~\ref{MainLemma}.

In this case the constant $C_3$ is given by:
\begin{align}\label{Cdefm'}
C_3 := \frac{2^{|E|} \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ R \min_{S \in E} L_k(\mathbf{AX}_{I(S)}) },
\end{align}
%
with $R$ being the lesser of $\min_{\substack{F \subseteq E}} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F})$ and $\min_{\substack{F \subseteq {[m'] \choose k}}} r( \{ \text{\rm Span}\{\mathbf{B}_{S}\} \}_{S \in F})$.
\end{proof}

>>>>

Since $f$ is a real analytic function, having one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set of (Borel) measure zero. Hence, almost every $n \times m$ real matrix $\mathbf{A}$ satisfies \eqref{SparkCondition} provided \eqref{CScondition} holds for a value of $\gamma$ for some distribution. We remark that the precise relationship between $m$, $n$, and $k$ guaranteeing that $f$ is not identically zero is a challenging problem in real algebraic geometry. In any case, we set $\gamma_0$ to be the smallest known such $\gamma$.

@@

%In fact, in the case $k=1$ we can relax our assumptions even further and assume that the matrix $\mathbf{B}$ has only at least as many columns as $\mathbf{A}$; as we will see, the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} then hold for some $n \times m$ submatrix of $\mathbf{B}$.  [*** ??? ***]

\begin{proof}[Proof ($k=1$)]
The only 1-uniform hypergraph satisfying the singleton intersection property is $[m]$; hence, we have $\mathbf{x}_i = c_i \mathbf{e}_i$ for $c_i \in \mathbb{R} \setminus \{\mathbf{0}\}$, $i \in [m]$. In this case we may take $C_1 = 1/ \min_{\ell \in [m]} |c_{\ell}|$.

Fix $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfying \eqref{SparkCondition} and suppose that for some $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ ($\hat m \geq m$) and $1$-sparse $\mathbf{\hat x}_i \in \mathbb{R}^{\hat m}$ we have  $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon < L_2(\mathbf{A}) / C_1$ for all $i$. Then there are $c'_1, \ldots, c'_m \in \mathbb{R}$ and $\pi: [m] \to [\hat m]$ with:
\begin{align}\label{1D}
\|c_j\mathbf{A}_j - c'_j\mathbf{B}_{\pi(j)}\|_2 \leq \varepsilon, \ \ \text{for all $j$}.
\end{align} 
Note that $c'_j \neq 0$ for all $j$ since otherwise by \eqref{delrho} and definition of $C_1$ we would have $\|c_j\mathbf{A}_j\|_2 < \min_{\ell \in [m]}\|c_{\ell}\mathbf{A}_{\ell}\|_2$.

We  now show that $\pi$ is injective (and thus is a permutation if $\hat m = m$). Suppose that $\pi(i) = \pi(j) = \ell$ for some $i \neq j$ and $\ell$. Then, $\|c_{j}\mathbf{A}_{j} - c'_{j}\mathbf{B}_{\ell}\|_2 \leq \varepsilon$ and $\|c_{i}\mathbf{A}_{i} - c'_{i} \mathbf{B}_{\ell}\|_2  \leq \varepsilon$. Scaling and summing these inequalities by $|c'_{i}|$ and $|c'_{j}|$, respectively, and applying the triangle inequality, we obtain:
\begin{align*}%\label{contra}
(|c'_{i}| + |c'_{j}|) \varepsilon
&\geq\|\mathbf{A}(c'_{i}c_{j} \mathbf{e}_{j} - c'_{j}c_{i}\mathbf{e}_{i})\|_2 \nonumber \\ 
&\geq  \left( |c'_{i}| + |c'_{j}| \right) L_2(\mathbf{A}) \min_{\ell \in [m]} |c_\ell |,
\end{align*}
%
which contradicts the bound $\varepsilon < L_2(\mathbf{A})/C_1$. Hence, the map $\pi$ is injective. Setting $J = \pi([m])$ and letting $P = \left( \mathbf{e}_{\pi(1)} \cdots \mathbf{e}_{\pi(m)}\right)$ and $D = \text{diag}(\frac{\hat{c}_1}{c_1},\ldots,\frac{\hat{c}_m}{c_m})$, we see that \eqref{1D} becomes, for all $j \in [m]$:
\begin{align*}%\label{k=1result}
\|\mathbf{A}_j - (\mathbf{B}_J\mathbf{PD})_j\|_2 
= \|\mathbf{A}_j - \frac{c'_j}{c_j}\mathbf{B}_{\pi(j)}\|_2 
\leq \frac{\varepsilon}{|c_j|} 
\leq C_1\varepsilon.
\end{align*}
\end{proof}

@@

\begin{definition}\label{dDef}
For any point $\mathbf{u}$, let $\text{\rm dist}(\mathbf{u}, V) := \inf \{\| \mathbf{u}-\mathbf{v} \|_2: \mathbf{v} \in V\}$ be the distance of $\mathbf{u}$ from the set $V$. For subspaces $U,V \subseteq \mathbb{R}^m$, we define: %[Kato p.197]
\begin{align}\label{d}
d(U,V) := \sup_{\mathbf{u} \in U, \ \|\mathbf{u}\|_2 = 1} \text{\rm dist}(\mathbf{u},V).
\end{align}
Note that $d(U,\{\textbf{0}\}) = 1$ for $U \neq \{\mathbf{0}\}$, whereas if $U = \{\textbf{0}\}$ then $d$ has no meaning; in this case, set $d(\{\textbf{0}\},V) = 0$ for any $V$.
\end{definition}
%Note that the supremum is always achievable by a point in $U$ when $U, V \subseteq \mathbb{R}^m$.

@@

Our result in combinatorial matrix analysis is the following, which we derive by the following arguments (see the Supplemental for the full proof). First, we show that the aforementioned proximity between $k$-dimensional subspaces implies a proximity between smaller subspaces spanned by columns of $\mathbf{A}$ indexed by the intersections of sets in $E$ and those spanned by as many or fewer columns of $\mathbf{B}$. Another pigeonholing argument here combined with our assumptions on $E$ (e.g., the singleton intersection property) reveals that, in fact, each column of $\mathbf{A}$ spans a subspace proximal to that spanned by some column of $\mathbf{B}$. \eqref{Cstable} is a simple consequence of this last fact, which actually constitutes the proof of the theorem for the case $k=1$. % We present this special case now before restating the above arguments in greater detail.

@@

[Effective computation of stable datasets, at least in principle.]
[maybe smooth analysis something here]

@@

\begin{lemma}\label{MainLemma2}
Fix integers $n$ and $k < m \leq \hat m$ and suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ and $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ both satisfiy \eqref{SparkCondition}. There exists a constant $C_4 > 0$ for which for all $\varepsilon < L_2(\mathbf{A}) / C_4$ the following holds. If for some $E \subseteq2^{[m]}$ satisfying the singleton intersection property there exists a size-preserving map $\pi: E \mapsto 2^{[\hat m]}$ satisfying:
\begin{align}\label{GapUpperBound}
d(\text{\rm Span}\{\mathbf{A}_{S}\}, \text{\rm Span}\{\mathbf{B}_{\pi(S)}\}) \leq \varepsilon \ \ \text{for all $S \in E$},
\end{align}
%
then for some $J \in {[\hat m] \choose m}$, permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$, we have:
\begin{align}\label{MainLemmaBPD}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq C_4 \varepsilon, \indent \text{for } j \in [m].
\end{align}
The constant $C_4$ is given by:
\begin{align}\label{Cdefm}
C_4 := \frac{2^{|E|}}{R} \max_{j \in [m]} \|\mathbf{A}_j\|_2
\end{align}
%
with $R$ being the lesser of $\min_{\substack{F \subseteq E}} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F})$ and $\min_{\substack{F \subseteq {[\hat m] \choose k}}} r( \{ \text{\rm Span}\{\mathbf{B}_{S}\} \}_{S \in F})$.
\end{lemma}

\begin{proof}
The proof is very similar to the case $\hat m = m$, the difference being that now we may not invoke Lem.~\ref{NonEmptyLemma} (which requires $\hat m = m$) to infer from \eqref{fact2} that $| \cap_{S \in F(i)} \pi(S) | = 1$ for all stars $F(i)$. We circumvent this issue by instead assuming the spark condition on $\mathbf{B}$, which allows us to swap the roles of $\mathbf{A}$ and $\mathbf{B}$ in the arguments leading to the derivation of \eqref{fact2} to prove the opposite inequality for every star $F(i)$. Of course, this requires that the constant $C_4$  depend on $\mathbf{B}$. By these arguments we establish that $| \cap_{S \in F(i)} \pi(S) | = 1$ for all stars $F(i)$, yielding an injective map $\hat \pi: [m] \to [\hat m]$. The result then follows from the proof of the case $k=1$. 
\end{proof}

@@

\begin{proof}[Proof of Thm.~\ref{DeterministicUniquenessTheorem2}]
The proof is the same as that of the case $\hat m = m$, only now we establish a map $\pi: E \to {[\hat m] \choose k}$ by pigeonholing $(k-1){\hat m \choose k} + 1$ vectors with respect to holes $[\hat m]$ and eventually applying Lem.~\ref{MainLemma2} instead of Lem.~\ref{MainLemma}.

In this case the constant $C_3$ is given by:
\begin{align}\label{Cdefm}
C_3 := \frac{2^{|E|} \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ R \min_{S \in E} L_k(\mathbf{AX}_{I(S)}) },
\end{align}
%
with $R$ being the lesser of $\min_{\substack{F \subseteq E}} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F})$ and $\min_{\substack{F \subseteq {[\hat m] \choose k}}} r( \{ \text{\rm Span}\{\mathbf{B}_{S}\} \}_{S \in F})$.
\end{proof}
